Lifting CDCL to Numerical Domains

### 2016-10-09 deductions.tex
introduction

\section{Deduction}
In a typical abstract interpretation based galois-connection settings over 
an over-approximate domain, every concrete elements have a unique over-approximate 
representation in the abstract.  Likewise, every concrete transformer is
over-approximated by a unique abstract transformer.  We now define a abstract 
deduction transformer.

@@

For a program with $N$ variables, let $L$ be the total number of 
meet irreducibles returned by a domain $D$.  For $D$ = {\em ItvDom}, the 
maximum value of $L$ is $2*N$, whereas the maximum value of $L$ is 
$2*(N^2)$ for $D$ = {\em OctDom}. Note that an octagon is the conjunction 
of all octagonal inequalities in the set $L$.
%
\begin{figure}[t]
\scriptsize
\begin{tabular}{l|l|l}
\hline
C program & SSA & Octagon \\
\hline
\begin{lstlisting}[mathescape=true,language=C]
int main() {
 unsigned x, y;
 __CPROVER_assume(x==y);
 x++;
 assert(x==y+1);
}
\end{lstlisting}
&
\begin{minipage}{4.40cm}
$\begin{array}{l@{\,\,}c@{\,\,}l}
SSA &\iff& ((guard0 == TRUE) \land \\
    &    & (cond == (x == y)) \land \\
    &    & (guard1 == (cond \&\& guard0)) \land \\
    &    & (x' == 1u + x) \land \\
    &    & (x' == 1u + y || !guard1))
\end{array}$
\end{minipage}
&
\begin{minipage}{3.75cm}
$\begin{array}{l@{\,\,}c@{\,\,}l}
C &\iff& ((x' > 1) \land (-x'-y < -2) \land \\
  &    & (-x-x' < -2) \land (y-x' < 0) \land \\                                                                
  &    & (x-x' < 0) \land (y > 0) \land \\
  &    & (x > 0) \land (-x'-y < 0) \land \\
  &    & (x+y > 1) \land (y-x < 1) \land \\
  &    & (x'-y < 2) \land (x-y < 1) \land \\
  &    & (x+y > 0) \land (x+x' > 0) \land \\
  &    & (x'-x < 2))
\end{array}$
\end{minipage}
\\
\hline
\end{tabular}
\caption{C Program, corresponding SSA and Octagonal Inequalities}
\label{swssa}
\end{figure}
%
%    
\begin{algorithm2e}[htbp]
\DontPrintSemicolon
\SetKwFunction{deduce}{deduce}
\SetKwFunction{print}{print}
\SetKwFunction{return}{return}
\SetKwFunction{continue}{continue}
\SetKwFunction{assign}{assign}
\SetKwData{sat}{SAT}
\SetKwData{unsat}{UNSAT}
\SetKwData{unknown}{UNKNOWN}
\SetKwData{true}{true}
\SetKwInOut{Input}{input}
\SetKwInOut{Output}{output}
\SetKwFor{Loop}{Loop}{}{}
\SetKw{KwNot}{not}
\begin{small}
\Input{Set $S$ of $\widehat{ded_{\varphi}}$ for $\varphi$ over Abstract domain $A$,
abstract value $v$, propagation trail $\mathcal{T}$, reason trail $\mathcal{R}$}
\Output{The status (\sat or \unsat or \unknown)}
worklist $\leftarrow$ initialize\_worklist($S$) \;
\Loop{}
{
// Greatest fixed point \;
\While{$!worklist.empty()$} 
{
  $\langle live\_vars, s \rangle \leftarrow$ worklist.pop() \;
  $\mathit{meet\_irrd} \leftarrow$ domain(s, live\_vars, $v$)\;
  // check if the domain deduce BOTTOM \;
  \uIf{$meet\_irrd == \bot$} {
    $\mathcal{R}[\bot] \leftarrow s$ \;
    // empty the worklist \;
    worklist.delete() \;
    \return \sat \;
  }
  \uElse
  {
    // add deduction to trail \;
    $\mathcal{T} \leftarrow \mathcal{T} . \mathit{meet\_irrd}$ \; 
    $v \leftarrow v \meet \mathit{meet\_irrd}$ \; 
    $\mathcal{R}[|\mathcal{T}|] \leftarrow s$ \;
    // update the worklist \;
    worklist.update(v, s) \; 
  }
}
// check for counterexample\;
\uIf{$v$ is $\gamma$-complete} {
  \return \unsat;
}
$d \leftarrow decide(v)$ \;
\uIf{$v \sqsubseteq d$} {
\return \unknown \;
}
// add decision to trail\;
$\mathcal{T} \leftarrow \mathcal{T} . d$ \; 
$v \leftarrow v \meet d$ \; 
$\mathcal{R}[|\mathcal{T}|] \leftarrow nil$ \;
}
\end{small}
\caption{Abstract Model Search\label{Alg:ms}}
\end{algorithm2e}
%
\subsection{Abstract Model Search}
Algorithm~\ref{Alg:ms} presents the abstract model search procedure. 
In propositional case, a model search involves two steps -- 
{\em deductions} using unit rule refines current partial assignments 
and {\em decisions} to heuristically guess a value for an unassigned 
literal.  A unit rule overapproximates a model transformer and deduction 
computes a greatest fixed point over the partial assignments
domain~\cite{dhk2013-popl}.  Similarly, an abstract model search computes
a greatest fixed point over meet irreducible deduction transformer in $S$.

A {\em worklist} is initialized following a heuristics choice for iteration 
strategies. A {\em forward} iteration strategy initializes the worklist with 
transformers that has constants in the right-hand side. Whereas, a {\em backward} 
iteration strategy starts with a set of assertions. On the other hand, a 
{\em chaotic} iteration strategy initializes the worklist with set of all 
transformers.  For every meet irreducible deduction transformer $s \in S$ 
passed to the abstract domain, if the domain infers a new meet irreducible 
$\mathit{meet\_irrd}$, it is added to $\mathcal{T}$.  Additionally, the reason 
trail $\mathcal{R}$ is updated with the transformer $s$ as the reason for inferring
$\mathit{meet\_irrd}$. If a conflict is deduced, that is, $\mathit{meet\_irrd} = \bot$ 
is inferred by the domain, the algorithm terminates with UNSAT.  However,
when a fixed-point is reached, the abstract valuation $v$ is checked for set of
models.  This is determined by checking whether the set of assignments in $v$ is 
$\gamma$-complete~\cite{gamma}.  This can be achieved by checking if there is a 
concrete solution $c$ such that the inequality $c(x) \subseteq \gamma \circ
v(x)$ holds for every meet irreducible $x$.  If $v$ is $\gamma$-complete, that is, the 
abstract valuation $v$ is abstractly satisfying, the abstract model 
search terminates with SAT.  Otherwise, a decision is made by calling a 
procedure $decide(v)$ which returns a meet irreducible $d$.  If $d$ refines the 
current abstract value $v$, then the corresponding propagation and reason trail,
$\mathcal{T}$ and $\mathcal{R}$, are updated. Else, the model search procedure 
terminates with UNKNOWN.

>>>>

\section{Abstract Model Search}
In a typical abstract interpretation based galois-connection settings over 
an over-approximate domain, every concrete elements have a unique over-approximate 
representation in the abstract.  Likewise, every concrete transformer is
over-approximated by a unique abstract transformer.  We now define a abstract 
deduction transformer.

@@

For a program with $N$ variables, let $L$ be the total number of 
meet irreducibles returned by a domain $D$.  For $D$ = {\em ItvDom}, the 
maximum value of $L$ is $2*N$, whereas the maximum value of $L$ is 
$2*(N^2)$ for $D$ = {\em OctDom}. Note that an octagon is the conjunction 
of all octagonal inequalities in the set $L$.
%
\begin{figure}[t]
\scriptsize
\begin{tabular}{l|l|l}
\hline
C program & SSA & Octagon \\
\hline
\begin{lstlisting}[mathescape=true,language=C]
int main() {
 unsigned x, y;
 __CPROVER_assume(x==y);
 x++;
 assert(x==y+1);
}
\end{lstlisting}
&
\begin{minipage}{4.40cm}
$\begin{array}{l@{\,\,}c@{\,\,}l}
SSA &\iff& ((guard0 == TRUE) \land \\
    &    & (cond == (x == y)) \land \\
    &    & (guard1 == (cond \&\& guard0)) \land \\
    &    & (x' == 1u + x) \land \\
    &    & (x' == 1u + y || !guard1))
\end{array}$
\end{minipage}
&
\begin{minipage}{3.75cm}
$\begin{array}{l@{\,\,}c@{\,\,}l}
C &\iff& ((x' > 1) \land (-x'-y < -2) \land \\
  &    & (-x-x' < -2) \land (y-x' < 0) \land \\                                                                
  &    & (x-x' < 0) \land (y > 0) \land \\
  &    & (x > 0) \land (-x'-y < 0) \land \\
  &    & (x+y > 1) \land (y-x < 1) \land \\
  &    & (x'-y < 2) \land (x-y < 1) \land \\
  &    & (x+y > 0) \land (x+x' > 0) \land \\
  &    & (x'-x < 2))
\end{array}$
\end{minipage}
\\
\hline
\end{tabular}
\caption{C Program, corresponding SSA and Octagonal Inequalities}
\label{swssa}
\end{figure}
%
%    
\begin{algorithm2e}[htbp]
\DontPrintSemicolon
\SetKwFunction{deduce}{deduce}
\SetKwFunction{print}{print}
\SetKwFunction{return}{return}
\SetKwFunction{continue}{continue}
\SetKwFunction{assign}{assign}
\SetKwData{sat}{SAT}
\SetKwData{unsat}{UNSAT}
\SetKwData{unknown}{UNKNOWN}
\SetKwData{true}{true}
\SetKwInOut{Input}{input}
\SetKwInOut{Output}{output}
\SetKwFor{Loop}{Loop}{}{}
\SetKw{KwNot}{not}
\begin{small}
\Input{Set $S$ of $\widehat{ded_{\varphi}}$ for $\varphi$ over Abstract domain $A$,
abstract value $v$, propagation trail $\mathcal{T}$, reason trail $\mathcal{R}$}
\Output{The status (\sat or \unsat or \unknown)}
worklist $\leftarrow$ initialize\_worklist($S$) \;
\Loop{}
{
// Greatest fixed point \;
\While{$!worklist.empty()$} 
{
  $\langle live\_vars, s \rangle \leftarrow$ worklist.pop() \;
  $\mathit{meet\_irrd} \leftarrow$ domain(s, live\_vars, $v$)\;
  \uIf{$meet\_irrd == \bot$} {
    $\mathcal{R}[\bot] \leftarrow s$ \;
    worklist.delete() \;
    \return \sat \;
  }
  \uElse
  {
    $\mathcal{T} \leftarrow \mathcal{T} . \mathit{meet\_irrd}$ \; 
    $v \leftarrow v \meet \mathit{meet\_irrd}$ \; 
    $\mathcal{R}[|\mathcal{T}|] \leftarrow s$ \;
    worklist.update(v, s) \; 
  }
}
\uIf{$v$ is $\gamma$-complete} {
  \return \unsat;
}
$d \leftarrow decide(v)$ \;
\uIf{$v \sqsubseteq d$} {
\return \unknown \;
}
$\mathcal{T} \leftarrow \mathcal{T} . d$ \; 
$v \leftarrow v \meet d$ \; 
$\mathcal{R}[|\mathcal{T}|] \leftarrow nil$ \;
worklist.update(v, d) \; 
}
\end{small}
\caption{Abstract Model Search\label{Alg:ms}}
\end{algorithm2e}
%
\subsection{Algorithm for Abstract Model Search}
Algorithm~\ref{Alg:ms} presents the abstract model search procedure. 
In propositional case, a model search involves two steps -- 
{\em deductions} using unit rule refines current partial assignments 
and {\em decisions} to heuristically guess a value for an unassigned 
literal.  A unit rule overapproximates a model transformer and deduction 
computes a greatest fixed point over the partial assignments
domain~\cite{dhk2013-popl}.  Similarly, an abstract model search computes
a greatest fixed point over meet irreducible deduction transformer in $S$.

A {\em worklist} is initialized following a heuristics choice for iteration 
strategies. A {\em forward} iteration strategy initializes the worklist with 
transformers that has constants in the right-hand side. Whereas, a {\em backward} 
iteration strategy starts with a set of assertions. On the other hand, a 
{\em chaotic} iteration strategy initializes the worklist with set of all 
transformers.  For every meet irreducible deduction transformer $s \in S$ 
passed to the abstract domain, if the domain infers a new meet irreducible 
$\mathit{meet\_irrd}$, it is added to $\mathcal{T}$.  Additionally, the reason 
trail $\mathcal{R}$ is updated with the transformer $s$ as the reason for inferring
$\mathit{meet\_irrd}$. If a conflict is deduced, that is, $\mathit{meet\_irrd} = \bot$ 
is inferred by the domain, the algorithm terminates with UNSAT.  However,
when a fixed-point is reached, the abstract valuation $v$ is checked for set of
models.  This is determined by checking whether the set of assignments in $v$ is 
$\gamma$-complete~\cite{gamma}.  This can be achieved by checking if there is a 
concrete solution $c$ such that the inequality $c(x) \subseteq \gamma \circ
v(x)$ holds for every meet irreducible $x$.  If $v$ is $\gamma$-complete, that is, the 
abstract valuation $v$ is abstractly satisfying, the abstract model 
search terminates with SAT.  Otherwise, a decision is made by calling a 
procedure $decide(v)$ which returns a meet irreducible $d$.  If $d$ refines the 
current abstract value $v$, then the worklist is updated with all transformers that are dependant on the current decision $d$ along with the update of the corresponding propagation and reason trail,
$\mathcal{T}$ and $\mathcal{R}$. Else, the model search procedure 
terminates with UNKNOWN.

### 2016-10-09 introduction.tex
definitions added
minor
introduction

\section{Introduction}
%

Most lattice structure used in static analysis have meet and join operations 
but lack negations. The meet operation precisely model conjunction, and
the join operation over-approximate disjunction. Thus, the precision loss in 
static analyzers is often due to imprecise join operations. The loss in precision is
overcome by using richer abstract domains or equipping the analysis with disjunction --
which is often very expensive since the analysis exhibit case enumeration behavior.

On the other hand, the combination of precise conjunction in partial assignments domain 
with learning allow CDCL solvers to reason about disjunction without 
enumerating cases. Learning can be viewed as synthesizing an abstract transformer for 
negation. ACDCL lifts learning techniques in SAT solver to operate on non-distributive 
lattice -- thus equipping the analysis with negation. This enables the analyzer to refine 
the analysis and prevent enumeration behavior.

>>>>

\section{Introduction}
%
A standard static analyses for program safety aims to compute 
invariants by applying transformers that refines an abstract 
value.  Most lattice structure used in static analysis have 
meet and join operations but lack negations.  The meet operation 
precisely model conjunction, and the join operation over-approximate 
disjunction. Thus, the precision loss in static analyzers is often 
due to imprecise join operations. The loss in precision is overcome 
by using richer abstract domains or equipping the analysis with disjunction --
which is often very expensive since the analysis exhibit case enumeration 
behavior.

On the other hand, the combination of precise conjunction in partial assignments 
domain with learning allow Conflict Driven Clause Learning (CDCL)~\cite{cdcl} 
solvers to reason about disjunction without enumerating cases.  
Silva et. al.~\cite{tacas12, sas12, dhk2013-popl} presented 
a formal mechanism for applying abstract interpretation to solve the 
satisfiability problem.  ~\cite{sas12} provide a mathematical generalization 
of CDCL algorithm called, {\em Abstract Conflict Driven Clause Learning} (ACDCL), 
in terms of deduction and abduction transformers, fixed-point, interpolation 
and extrapolation operators on lattices.  The partial assignments in the 
CDCL algorithm are an abstract domain, unit rule is the best abstract 
transformer,  Boolean Constraint Propagation (BCP) computes a greatest fixed-point 
over the abstract domain, decision is dual widening and clause learning can be 
viewed as synthesizing an abstract transformer for negation.  Thus, ACDCL 
lifts learning techniques in SAT solver to operate on non-distributive 
lattice -- thus equipping the analysis with negation. This enables the 
analyzer to refine the analysis and prevent enumeration behavior.

The instantiation of ACDCL has so far been limited to interval domain only.  In this paper, 
we identify specific properties of the CDCL algorithm that are necessary to 
lift the propositional skeleton of CDCL to other richer lattice structures 
and relational abstract domains such as octagons, polyhedra, zones and equalities.  
To this end, we present a characterization of abstract model search procedure and 
conflict analysis procedure that lift the abstract deduction transformer and the 
first-UIP algorithm to relational lattice structures.
